---
title: "Get Started with facet"
author:
  - name: George Michaelides
    affiliation: University of East Anglia
  - name: Duncan J.R. Jackson
    affiliation: King's College London
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started with facet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The `facet` package provides a unified framework for Generalizability Theory (G-Theory). In this quick start, we will walk through a standard analysis: conducting a G-study, and then running a D-study.

## 1. Load the Package and Data

First, load `facet` and one of the built-in datasets, `brennan`.

```{r setup}
library(facet)
data(brennan)
head(brennan)
```

**Reading the data.** The `brennan` dataset has one row per observation: 10 persons × 3 tasks × 4 raters per task = 120 rows. Each row records one rater's score for one person on one task. This is a *nested* design: the same four raters do not rate every task; rather, each task has its own panel of raters. Because the `Rater` factor in the data has 12 uniquely labelled levels (4 per task), `(1 | Rater)` and `(1 | Rater:Task)` are mathematically equivalent random-effect specifications; the canonical form uses `(1 | Rater)`. The `Score` column is the observed rating; the other three columns identify which cell of the design the row belongs to.

## 2. Conduct a G-Study

Use the `gstudy()` function to estimate variance components. By default, `facet` uses `lme4` (Restricted Maximum Likelihood) under the hood. The canonical "all possible variance components" formula for the brennan data includes the three main effects (Person, Task, Rater) plus the only estimable interaction (Person:Task); the Person × Rater:Task interaction is confounded with the residual because there is one observation per cell.

```{r gstudy}
g_obj <- gstudy(Score ~ (1 | Person) + (1 | Task) + (1 | Rater) +
  (1 | Person:Task),
  data = brennan)

# View the variance components
print(g_obj)
```

**Interpreting the variance components.** The output shows four variance components: `Person` (universe score variance — true differences between persons), `Task` (differences in average difficulty across tasks), `Rater` (differences in leniency/severity between raters, which here are within-task rater effects because the 12 Rater labels are unique to each task), and `Person:Task` (the interaction — whether a person performs better on some tasks than others). The "pct" column shows what fraction of total variance each component accounts for. A healthy measurement has a large `Person` component (typically >30%) and a small `Person:Task` interaction (the test is fair across tasks). If a facet is taking up a large fraction of variance, that is the facet to invest more levels in (more tasks, more raters) for your next measurement campaign.

## 3. Visualize Variance Components

Plotting the random effects helps identify which facets contribute the most variance.

```{r plot}
plot(g_obj)
```

**Reading the variance plot.** Each bar represents a variance component, and the height shows its magnitude (on the variance or proportion scale depending on the `type` argument). The most useful patterns to look for: (1) a tall `Person` bar means your instrument is good at distinguishing individuals; (2) a tall facet bar (e.g., `Task` or `Rater:Task`) means that facet is a major source of error and you should add more levels of it; (3) similar heights across components suggest a balanced design where no single facet dominates. Compare the bar heights to the "pct" column from step 2 to confirm the relative contributions.

## 4. Conduct a D-Study

A Decision (D) study calculates the dependability of your measurement design. You can project how changing the sample sizes of your facets impacts reliability.

```{r dstudy}
# Project a design with 3 Tasks and 4 Raters
d_obj <- dstudy(g_obj, n = list(Task = 3, Rater = 4))

print(d_obj)
```

**Interpreting the D-study output.** Two coefficients are reported: **G** (generalizability coefficient, for relative decisions like "rank these persons") and **Phi** (dependability coefficient, for absolute decisions like "did this person meet the standard?"). Conventional thresholds: G ≥ 0.80 for high-stakes decisions (selection, certification), G ≥ 0.70 for medium-stakes (formative feedback, progress monitoring), G ≥ 0.60 for low-stakes research. Phi is always ≤ G; the gap between them tells you how much absolute-error variance (e.g., task-difficulty shifts) is contributing. If the projected G is below your target, the next step is to run a sweep — vary `n` across multiple values to find the smallest design that achieves the reliability you need.

## Next Steps

For comprehensive documentation, see `vignette("introduction")`, which covers:

- Historical foundations of generalizability theory
- All three estimation estimators (aov, REML, Bayesian)
- Univariate and multivariate G-studies and D-studies
- Value Added Ratios and composite score optimization
- Advanced topics, reporting guidelines, and exercises
