---
title: "Using facet for Generalizability Theory"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using facet for Generalizability Theory}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction to facet

The **facet** package provides comprehensive tools for conducting Generalizability Theory (G-theory) analyses using variance component models. It supports three estimation backends: frequentist (lme4), Bayesian (brms), and method of moments (mom/ANOVA-based). The package provides a unified interface for univariate and multivariate designs, with full support for G-studies, D-studies, and coefficient calculations with uncertainty quantification.

**Note**: The `tidy()` and `glance()` methods require the broom package to be loaded. The examples below use `library(broom)` to ensure these functions are available.

## Overview

Generalizability Theory provides a framework for quantifying the reliability of measurements across multiple sources of error. The key concepts are:

- **G-studies**: Estimate variance components associated with different facets of measurement (e.g., items, raters, occasions)
- **D-studies**: Compute generalizability (G) and dependability (D/Phi) coefficients for different measurement designs

The facet package supports multiple backends:

- **lme4**: For frequentist estimation using linear mixed effects models (default for univariate models)
- **brms**: For Bayesian estimation using Stan (required for multivariate models)
- **mom**: For method of moments (ANOVA-based) estimation. Works best with balanced designs and provides confidence intervals using asymptotic approximations.

## Installation

```r
# Install from GitHub
# install.packages("devtools")
devtools::install_github("yourusername/facet")
```

## Basic G-Study

A G-study estimates the variance components for each facet in your measurement design. The formula syntax follows **lme4** conventions, specifying random effects with `(1 | facet)`.

In this example, we use the `brennan` dataset, which contains scores for 10 persons across 3 tasks and 12 raters in a fully crossed design ($p \times t \times r$).

```{r}
library(facet)
library(dplyr)
library(tidyr)
library(broom)

# Load Brennan dataset
data(brennan)

# Conduct G-study with default (lme4) backend
# We specify a person-by-task design.
g <- gstudy(Score ~ (1 | Person) + (1 | Task), 
            data = brennan)

# View variance components
print(g)
```

The output shows the estimated variance components for each facet:

- **person**: Variance attributable to differences between persons (universe score variance)
- **rater**: Variance attributable to differences between raters
- **Residual**: Remaining variance (person × rater interaction + error)

### Viewing Results

Use `summary()` for a more detailed view, with options to display estimates on variance or SD scale:

```{r}
# Summary with variance estimates (default)
summary(g)

# Summary with standard deviation estimates
summary(g, scale = "sd")
```

Extract the variance components as a tibble using `tidy()`:

```{r}
tidy(g)
```

Use `glance()` for a one-row summary of the G-study:

```{r}
glance(g)
```

## G-Study with Confidence Intervals

For the **lme4** backend, you can request confidence intervals using the `ci_method` parameter:

- `"none"` (default): No confidence intervals
- `"profile"`: Profile likelihood confidence intervals (more accurate, slower)
- `"boot"`: Parametric bootstrap confidence intervals (most accurate, slowest)

```{r}
# G-study with profile likelihood confidence intervals
g_profile <- gstudy(
  Score ~ (1 | Person) + (1 | Task),
  data = brennan,
  ci_method = "profile"
)

summary(g_profile)
```

For bootstrap confidence intervals, you can specify the bootstrap type and number of simulations:

```{r}
# G-study with bootstrap confidence intervals
# Using fewer iterations for speed in this example
g_boot <- gstudy(
  Score ~ (1 | Person) + (1 | Task),
  data = brennan,
  ci_method = "boot",
  nsim = 50,  # Number of bootstrap samples
  boot.type = "perc"  # Percentile bootstrap
)

summary(g_boot)
```

## Method of Moments Backend

The method of moments (MOM) backend uses ANOVA-based variance component estimation. This approach works best with balanced designs and provides confidence intervals using asymptotic approximations.

```{r}
# Method of moments G-study
g_mom <- gstudy(
  Score ~ (1 | Person) + (1 | Task) + (1 | Rater) + 
          (1 | Person:Task),
  data = brennan,
  backend = "mom"
)

# Summary shows confidence intervals
summary(g_mom)
```

The MOM backend automatically handles complex designs and nested data correctly if specified in the formula.

```{r}
# Example with nested data (Method of Moments)
# Using rajaratnam dataset: items nested within subtests
data(rajaratnam)

g_nested <- gstudy(
  Score ~ (1 | Person) + (1 | Subtest) + (1 | Item:Subtest) + (1 | Person:Subtest),
  data = rajaratnam,
  backend = "mom"
)

print(g_nested)
```

## Bayesian Estimation with brms

For Bayesian estimation, use the **brms** backend. This provides:

- Posterior distributions for variance components
- Credible intervals (highest posterior density intervals)
- Full Bayesian inference including convergence diagnostics

```{r, eval=FALSE}
# Bayesian G-study with brms
# Note: Requires brms and Stan to be installed
g_bayes <- gstudy(
  Score ~ (1 | Person) + (1 | Task),
  data = brennan,
  backend = "brms"
)

print(g_bayes)
summary(g_bayes)
```

The **brms** backend provides additional diagnostic information like Rhat (Gelman-Rubin diagnostic) and Effective Sample Size (ESS) values.

## D-Study

A D-study uses the G-study results to compute reliability coefficients for different measurement designs. The key coefficients are:

- **G coefficient**: Generalizability coefficient for *relative* decisions (comparing individuals' rankings)
- **D coefficient**: Dependability ($\Phi$) coefficient for *absolute* decisions (comparing to a standard)

### Basic D-Study

```{r}
# Conduct D-study with 3 tasks and 4 raters
# (using our previously fitted lme4 g-study 'g')
d <- dstudy(g, n = list(Task = 3, Rater = 4))

print(d)
summary(d)
```

### Exploring Different Sample Sizes

You can explore multiple sample size combinations using a sweep:

```{r}
# Explore different numbers of tasks and raters
d_sweep <- dstudy(g, n = list(Task = c(3, 5, 10), Rater = c(2, 4, 8)))

# View coefficients for each design
# Using tidy() to get a nice data frame of coefficients
library(broom)
tidy(d_sweep)
```

## Specifying Object of Measurement

By default, the first facet in the formula is considered the object of measurement (typically "Person"). You can specify this explicitly or modify the universe/error structure:

```{r}
# D-study with custom universe specification
# The object of measurement is Person
d_custom <- dstudy(g, n = list(Task = 3, Rater = 4), 
                   universe = c("Person", "Person:Task"))
```

The `universe` parameter accepts multiple formats:

- Character string: `"Person"`
- Character vector: `c("Person", "Person:Task")`
- Formula: `~ Person + Person:Task`

Note: The object of measurement (first facet from the G-study) is always included in the universe. If you specify a universe without including the object, a warning is issued and the object is added automatically.

The universe score variance is calculated as follows:
- The object component is **not scaled**.
- Non-object universe components are scaled by their corresponding sample sizes.
- For example, if the object is "Person" and universe includes "Person:Task", then:
  - $\sigma^2(\text{Universe}) = \sigma^2(P) + \sigma^2(PT) / n_t$

## Advanced D-Study: Custom Error Specification

You can customize which variance components contribute to the error term using the `error` parameter:

```{r}
# D-study with custom error specification
# Using only interaction terms as error sources
d_error <- dstudy(
  g,
  n = list(Task = 3, Rater = 4),
  error = c("Person:Task", "Person:Rater")
)

summary(d_error)
```

## D-Study with Aggregation

When aggregating over a facet (e.g., taking the mean across multiple raters), use the `aggregation` parameter to correctly scale the variance components:

```{r}
# D-study with aggregation over raters
# This models the case where we average scores across Tasks or Raters
d_agg <- dstudy(
  g,
  n = list(Task = 3, Rater = 4),
  aggregation = "Rater",
  residual_is = "Person:Task:Rater"
)

summary(d_agg)
```

The `residual_is` parameter specifies which facets make up the residual variance component (important for proper scaling when aggregating).

## Posterior-Based D-Study Coefficients

For Bayesian G-studies, you can compute D-study coefficients using the full posterior distributions:

```{r, eval=FALSE}
# D-study with posterior estimation (requires brms backend)
# Note: This uses full posterior to compute coefficients with uncertainty
d_post <- dstudy(
  g_bayes,
  n = list(rater = 3),
  estimation = "posterior"
)

print(d_post)

# Access posterior distributions
names(d_post$posterior)
```

If you request posterior estimation but the G-study was not fit with brms, the function will automatically refit:

```{r, eval=FALSE}
# D-study with posterior from lme4 gstudy
d_post2 <- dstudy(
  g,
  n = list(rater = 3),
  estimation = "posterior"
)
```

## Visualization

The **facet** package provides built-in plotting methods using **ggplot2**.

### G-Study Plots

```{r}
# Bar plot of variance components (proportions)
plot(g, type = "proportion")
```

```{r}
# Bar plot on variance scale
plot(g, type = "variance")
```

### D-Study Plots

```{r}
# Bar plot of G and D coefficients
plot(d, type = "coefficients")
```

```{r}
# Sweep plot (for sample size exploration)
plot(d_sweep, type = "sweep")
```

### Caterpillar Plots for Random Effects

Create caterpillar plots (dot plots with error bars) to display random effects and their uncertainty using `plot_ranef()`:

```{r}
# Plot all random effects
plot_ranef(g)

# Plot a specific facet
plot_ranef(g, which = "Person")
```

## Working with Results

### Extract Variance Components

```{r}
# Get variance components as data frame
vc <- tidy(g)
vc
```

### Filter and Manipulate

```{r}
# Filter specific components
vc %>%
  filter(component != "Residual") %>%
  mutate(pct_total = var / sum(var) * 100)
```

### Access Model Objects

The underlying model object can be accessed from a gstudy object:

```{r}
# Access the fitted model
model <- g$model
print(summary(model))
```

**Note**: The `VarCorr()` and `ranef()` functions work with the underlying model when lme4 is loaded:

```{r, eval=FALSE}
# Extract variance-covariance matrices (requires loading lme4)
library(lme4)
vc_obj <- VarCorr(g)
print(vc_obj)

# Extract random effects (BLUPs/conditional modes)
re <- ranef(g)
print(re)
```

For brms models:

```{r, eval=FALSE}
# Extract random effects from Bayesian model
re_bayes <- ranef(g_bayes)
print(re_bayes)

# Parameter pairs plot
# pairs(g_bayes)
```

### Extract Posterior Draws

For brms backend, extract posterior draws for custom analyses:

```{r, eval=FALSE}
# Extract posterior draws
draws <- brms::as_draws_matrix(g_bayes$model)
```

## Multivariate Analysis

For multivariate designs (multiple outcome variables), use the **brms** backend syntax with `mvbind()`.

```{r, eval=FALSE}
# Create multivariate data by spreading the tasks
brennan_wide <- brennan %>%
  pivot_wider(names_from = Task, values_from = Score, names_prefix = "Task")

# Multivariate G-study (automatically uses brms)
g_mv <- gstudy(
  mvbind(Task1, Task2, Task3) ~ (1 | Person) + (1 | Rater),
  data = brennan_wide
)

summary(g_mv)

# D-study for multivariate models
d_mv <- dstudy(g_mv, n = list(Rater = 4))
print(d_mv)
```

## Three-Facet Designs

A common design in performance assessment includes person, task, and rater (already demonstrated with our G-study `g`):

```{r}
# G-study with three facets (see Basic G-Study section)
# D-study: what if we used 5 tasks rated by 2 raters?
d_3f <- dstudy(
  g,
  n = list(Task = 5, Rater = 2)
)

summary(d_3f)
```

## Complete Example

Here's a complete workflow demonstrating multiple features using the `brennan` dataset:

```{r}
library(facet)
library(dplyr)

# 1. Load data
data(brennan)

# 2. Conduct G-study with profile confidence intervals
g <- gstudy(
  Score ~ (1 | Person) + (1 | Task) + (1 | Rater) +
          (1 | Person:Task),
  data = brennan,
  ci_method = "profile"
)

# 3. View results
summary(g)

# 4. Conduct D-study with specific design (e.g., 5 tasks, 3 raters)
d <- dstudy(g, n = list(Task = 5, Rater = 3))

# 5. View coefficients
summary(d)

# 6. Explore alternative designs (Sample Size Sweep)
d_alt <- dstudy(g, n = list(Task = c(2, 5, 10), Rater = c(1, 2, 5)))

# 7. Plot sweep results
plot(d_alt, type = "sweep")

# 8. Extract and analyze variance components
tidy(g) %>%
  filter(component != "Residual") %>%
  arrange(desc(var))
```

## Utility Functions

### Predicates

Check object types:

```{r}
is.gstudy(g)
is.dstudy(d)
```

### Scale Parameter

Control display scale in print and summary methods:

```{r}
# Display on SD scale
print(g, scale = "sd")
summary(g, scale = "sd")
```

## References

- Brennan, R. L. (2001). *Generalizability theory*. Springer.
- Shavelson, R. J., & Webb, N. M. (1991). *Generalizability theory: A primer*. Sage Publications.
- Webb, N. M., Shavelson, R. J., & Haertel, E. H. (2006). 6 Reliability Coefficients and Generalizability Theory. *Handbook of statistics*, 26, 81-124.
